\documentclass[4pt,a4paper]{report}
\input{latex/preamble.tex}

\title{Toward a Formalization of 'Virtual Sets'}
\author{Christina O'Donnell}
\subtitle{School of Computer Science, University of Nottingham}
\date{September 2025}
\begin{document}

\maketitle

\section*{Acknowledgements}
I would like to thank Thorsten Altenkirch
for helping direct this project and for his invaluable feedback. I
also must thank Mark Williams for sharing his idea with me that
motivated dissertation.

\section*{Delcaration}
Submitted 21st September 2025, in partial fulfilment of the condiations for the award of the degree of MSc (Computer Science).

\begin{abstract}
  This dissertation investigates the categorical structure underlying
  the monoidal category of finite sets and injective functions, with
  the goal of understanding whether it admits a well-defined trace
  operator. A trace is a categorical generalization of feedback
  structures such as recursion, or feedback, introduced by Joyal
  et. al. (1996), traces can be applied to perform the role of
  recursion in categorical semantics of programming langauges, and the
  role of feedback in quantum computing. Mark Williams (2025) recently
  conjectured that the category of finite sets and injective functions
  may have a sufficient trace operator, however the construction has
  yet to be formally verified.

  The notion of 'virtual set', introduced by Williams (2025) is a set
  that can contain 'negative elements', satisfying the condition for a
  compact closed category. This construction amounts to taking a
  quotient of $\mathrm{Inj}^2$, and then proving coherence of the
  result, as spelled out in Joyal et. al. (1996). The difficulty is to
  prove all of the coherence lemmas required to show it has monoidal,
  and trace structure, which is independent of equivalent pairs of
  functions. Such a structure is of interest in simplicial
  type theory, where all inclusion map between simplicial complexes
  must necessarily be injective.

  This dissertation proceeds by developing the category \textrm{Inj}
  of finite sets and injective functions, constructing a composition
  operator and proving associativity and unit laws. A tensor product
  is then defined as well as a partial proof of the monoidal laws.
  We then define a trace operator and prove some of the required
  properties about its behavior. Finally, the \textrm{Int}
  construction is discussed.

  All definitions and proofs are implemented in cubical Agda, making
  this one of the first attempts to formalize the notion of traced
  monoidal categories. We discuss my two main approaches at
  formalizing the monoidal category of injective functions, and
  constructing the tensor operator, and trace, as well as proving some
  of the properties about it. \\
  \textbf{Keywords}: Trace Monoidal Categories, Compact Closed Categories, Cubical Agda 
\end{abstract}
\tableofcontents
\listoffigures

\include{latex/chapters/introduction}
\include{latex/chapters/prelim}
\include{latex/chapters/lit-review}
\include{latex/chapters/formalization}
\include{latex/chapters/conclusion}

\appendix
\chapter{References}
\bibliographystyle{plain}
\bibliography{latex/report}
\chapter{Agda Source Listings}
\end{document}
